---
title: "Success Probability Contour Report"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
    number_sections: true
    extra_dependencies: ["float"]
params:
  alpha: 2
  K: 10
  n: 10000
  B: 500
  cutoff: 0.05
  ae_grid_n: 60
  are_grid_n: 60
  seed: 260926
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      fig.width = 7, fig.height = 5)
knitr::opts_chunk$set(fig.pos = "H")

knitr::opts_knit$set(root.dir = here::here())
source(here::here("scripts", "simulations", "simulation_success_surface.R"))
```

# Overview

This report estimates and visualises success probabilities over a 2D grid of
AE and ARE thresholds for a single hybrid cutoff value \(c\).

**Success rule** — for each cell type \(k\):

- If the *true* proportion \(p_k < c\), success requires \(\text{AE}_k \le \tau_{AE}\).
- If \(p_k \ge c\), success requires \(\text{ARE}_k \le \tau_{ARE}\).

A simulation replicate is a success only when *all* \(K\) cell types
simultaneously satisfy their respective condition.

```{r configuration}
config <- list(
  alpha    = as.numeric(params$alpha),
  K        = as.integer(params$K),
  n        = as.integer(params$n),
  B        = as.integer(params$B),
  cutoff   = as.numeric(params$cutoff),
  ae_grid  = seq(0.001, 0.10,  length.out = as.integer(params$ae_grid_n)),
  are_grid = seq(0.01,  2.0,   length.out = as.integer(params$are_grid_n)),
  seed     = as.integer(params$seed)
)

setup_table <- data.frame(
  Parameter = c("alpha", "K", "n", "B", "cutoff",
                "AE grid", "ARE grid", "seed"),
  Value = c(
    config$alpha,
    config$K,
    config$n,
    config$B,
    config$cutoff,
    paste0("seq(0.001, 0.10, length.out = ", params$ae_grid_n, ")"),
    paste0("seq(0.01, 2.0, length.out = ",   params$are_grid_n, ")"),
    config$seed
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  setup_table,
  caption = "Simulation configuration used in this report.",
  booktabs = TRUE
)
```

```{r run-simulation}
result <- run_success_surface(config)
```

\newpage

# True proportions

```{r proportion-table}
p_df <- data.frame(
  index = seq_along(result$p),
  p     = round(result$p, digits = 5)
)
knitr::kable(
  p_df,
  caption = paste0(
    "True proportions for alpha = ", config$alpha,
    ". Cell types with p < ", config$cutoff, " use AE; others use ARE."
  ),
  booktabs = TRUE
)
```

\newpage

# Success probability contour plot

The contour lines mark estimated success probabilities from
`seq(0.1, 0.9, 0.1)`.  The x-axis is the AE threshold and the y-axis is the
ARE threshold.  Axis limits are set to `[0.001, 0.05]` (AE) and
`[0.01, 1]` (ARE); data outside these windows contribute to contour
estimation but are not displayed.

```{r contour-plot, fig.cap=paste0("Success probability contours for cutoff = ", config$cutoff, ".")}
print(result$contour_plot)
```
