---
title: "Simulation Error-Choice Report"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
params:
  alpha: [2, 2.5, 3, 4, 5]
  K: 10
  n: 1000
  B: 5000
  seed: 260926
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 10, fig.height = 5)

source(file.path("scripts", "simulation_functions.R"))
source(file.path("scripts", "simulation_errorchoice.R"))
```

# Overview

This report evaluates multinomial sampling error across cell-type proportions
generated from the normalized Beta(\(\alpha, 1\)) construction.

For each \(\alpha\), the workflow:

1. Generates true proportions over \(K\) cell types.
2. Simulates multinomial count draws.
3. Computes maximum error across cell types for Absolute Error (AE) and Absolute Relative Error (ARE).
4. Summarizes threshold-based success rates and the cell-type index most frequently driving the maximum error.

# Simulation setup

```{r configuration}
base_config <- simulation_errorchoice_defaults()
taus <- base_config$taus

config <- list(
  alpha = as.numeric(params$alpha),
  K = as.integer(params$K),
  n = as.integer(params$n),
  B = as.integer(params$B),
  taus = taus,
  metrics = c("AE", "ARE"),
  model = "multinomial",
  tie_method = "random",
  seed = as.integer(params$seed)
)

setup_table <- data.frame(
  item = c("alpha", "K", "n", "B", "metrics", "model", "tie_method", "seed"),
  value = c(
    paste(config$alpha, collapse = ", "),
    config$K,
    config$n,
    config$B,
    paste(config$metrics, collapse = ", "),
    config$model,
    config$tie_method,
    config$seed
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  setup_table,
  col.names = c("Parameter", "Value"),
  caption = "Simulation configuration used in this report."
)
```

```{r run-simulation}
result <- run_simulation_errorchoice(config)
```

# True proportions from Beta(\(\alpha, 1\))

## Proportion curves

```{r proportion-curves, fig.cap="Normalized Beta(alpha,1) curves and grid-point proportions by alpha."}
generate_proportions_curve(alpha = config$alpha, K = config$K)
```

## Numeric proportions per alpha

```{r proportion-table}
p_table <- result$p_table
value_cols <- grep("^index_", names(p_table), value = TRUE)
p_table[value_cols] <- lapply(p_table[value_cols], function(x) signif(x, 6))
knitr::kable(p_table, caption = "True proportion values by alpha and cell-type index.")
```

# Success-rate curves

## AE and ARE comparison

```{r success-rate-curves-both, fig.cap="Success-rate curves by alpha, shown separately for AE and ARE."}
plot_success_rate_curve(result, metric = NULL)
```

## AE-only

```{r success-rate-curves-ae, fig.cap="AE success-rate curves across alpha values."}
plot_success_rate_curve(result, metric = "AE")
```

## ARE-only

```{r success-rate-curves-are, fig.cap="ARE success-rate curves across alpha values."}
plot_success_rate_curve(result, metric = "ARE")
```

# Argmax histograms

## AE and ARE comparison

```{r argmax-hist-both, fig.cap="Argmax histograms by alpha and metric for maximum-error indices."}
plot_argmax_histogram(result, metric = NULL)
```

## AE-only

```{r argmax-hist-ae, fig.cap="Argmax histogram for AE across alpha values."}
plot_argmax_histogram(result, metric = "AE")
```

## ARE-only

```{r argmax-hist-are, fig.cap="Argmax histogram for ARE across alpha values."}
plot_argmax_histogram(result, metric = "ARE")
```

# Argmax summary tables

```{r argmax-summary}
argmax_summary <- result$argmax_summary
argmax_summary$fraction <- signif(argmax_summary$fraction, 4)
argmax_summary$p_value <- signif(argmax_summary$p_value, 6)
knitr::kable(argmax_summary, caption = "Argmax summary per alpha and error metric.")
```
