---
title: "Simulation Hybrid Cutoff Heatmaps Report"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: false
    number_sections: true
    extra_dependencies: ["float"]
params:
  alpha:
    value: [2, 2.5, 3, 4, 5]
  K: 10
  n: 10000
  B: 500
  cutoffs_by: 0.01
  tau_AE_values:
    value: [0.005, 0.01, 0.02, 0.03, 0.04]
  tau_ARE_values:
    value: [0.025, 0.05, 0.1, 0.2, 0.5]
  maximize: "cell"
  seed: 260926
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.width = 7, fig.height = 4)
knitr::opts_chunk$set(fig.pos = "H")

knitr::opts_knit$set(root.dir = here::here())
source(here::here("scripts", "simulation_functions.R"))
source(here::here("scripts", "simulation_hybrid_cutoff_heatmaps.R"))
```

# Overview

This report shows hybrid AE/ARE cutoff heatmaps for each of the five \(\alpha\) values.

For a given pair of thresholds (\(\tau_{AE}\), \(\tau_{ARE}\)), the best hybrid cutoff \(c^*\) is the value of \(c\) that maximises the success rate under the rule: use AE when \(\hat{p} \leq c\), otherwise use ARE.

The simulation uses the following settings:

```{r configuration}
config <- list(
  alpha          = as.numeric(params$alpha),
  K              = as.integer(params$K),
  n              = as.integer(params$n),
  B              = as.integer(params$B),
  cutoffs        = seq(0, 0.5, by = params$cutoffs_by),
  tau_AE_values  = as.numeric(params$tau_AE_values),
  tau_ARE_values = as.numeric(params$tau_ARE_values),
  maximize       = params$maximize,
  seed           = as.integer(params$seed)
)

setup_table <- data.frame(
  Parameter = c("alpha", "K", "n", "B", "cutoffs", "tau_AE values", "tau_ARE values",
                "maximize", "seed"),
  Value = c(
    paste(config$alpha, collapse = ", "),
    config$K,
    config$n,
    config$B,
    paste0("seq(0, 0.5, by = ", params$cutoffs_by, ")"),
    paste(config$tau_AE_values, collapse = ", "),
    paste(config$tau_ARE_values, collapse = ", "),
    config$maximize,
    config$seed
  ),
  stringsAsFactors = FALSE
)

knitr::kable(
  setup_table,
  caption = "Simulation configuration used in this report.",
  booktabs = TRUE
)
```

```{r run-simulation}
result <- run_simulation_hybrid_heatmaps(config)
```

\newpage

# True proportions

The true proportions for each \(\alpha\) are shown below. Proportions are derived from the weighted Beta(\(\alpha\), 1) quantile grid.

```{r proportion-table}
p_tab <- round(result$p_table, digits = 5)
p_tab <- p_tab[, !names(p_tab) %in% "p_max", drop = FALSE]
knitr::kable(p_tab, caption = "True proportion values by alpha and cell-type index.")
```

\newpage

# Hybrid cutoff heatmaps

Each heatmap shows, for a grid of (\(\tau_{AE}\), \(\tau_{ARE}\)) threshold pairs, the best hybrid cutoff \(c^*\). Cell labels give the cutoff value; colour intensity indicates magnitude.

```{r heatmap-alpha-1, fig.cap=paste0("Best hybrid cutoff heatmap for alpha = ", config$alpha[1], ".")}
print(result$heatmaps[[1]])
```

```{r heatmap-alpha-2, fig.cap=paste0("Best hybrid cutoff heatmap for alpha = ", config$alpha[2], ".")}
print(result$heatmaps[[2]])
```

```{r heatmap-alpha-3, fig.cap=paste0("Best hybrid cutoff heatmap for alpha = ", config$alpha[3], ".")}
print(result$heatmaps[[3]])
```

```{r heatmap-alpha-4, fig.cap=paste0("Best hybrid cutoff heatmap for alpha = ", config$alpha[4], ".")}
print(result$heatmaps[[4]])
```

```{r heatmap-alpha-5, fig.cap=paste0("Best hybrid cutoff heatmap for alpha = ", config$alpha[5], ".")}
print(result$heatmaps[[5]])
```
